---
title: "Lymphocyte Data Diagnosys Script"
author: "Giménez Gredilla, Daniel"
date: "November 2017"
bibliography: library.bib
output:
  pdf_document: 
    toc: yes
    toc_depth: 3
    fig_caption: yes
    includes:  
      in_header: tex-support.tex
options:
  encoding: utf8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.lp = "Fig. ", tidy.opts=list(width.cutoff=60),tidy=TRUE, cache = TRUE, fig.width=12, fig.height=4, comment = NULL, out.height='200px', dpi=200, fig.pos = 'h', out.extra = '')
```

```{r library_load, include=FALSE, echo=FALSE, warning=FALSE}
require(car)
require(psych)
require(caret)
require(mlbench)
require(randomForest)
```

```{r data_var_declare, include=FALSE, echo=FALSE}
data_class_factors <- 7
data_class_numeric <- 2867
data_response_index <- 2
```

```{r data_load, include=FALSE, echo=FALSE}
colClasses <- append(c(rep("factor",data_class_factors)),c(rep("numeric",data_class_numeric)))
#data_df <- read.csv2(file = "resources/data_dummy.csv", sep = ",", dec = ".", colClasses = colClasses, stringsAsFactors = FALSE)
data_df <- read.csv2(file = "resources/data.csv", sep = ",", dec = ".", colClasses = colClasses, stringsAsFactors = FALSE)
data_formula <- as.formula(paste(colnames(data_df[data_response_index]), "~", paste(colnames(data_df)[-(1:data_class_factors)], collapse=" + ")))
```

```{r data_df_inf, include=FALSE, echo=FALSE}
#data_control <- rfeControl(functions=rfFuncs, method="cv", number=10)
#data_rfe <- rfe(data_df[,(data_class_factors+1):length(data_df[1,])], data_df[,data_response_index], sizes = c(1:20), metric = "Accuracy", maximize = TRUE, rfeControl = data_control)
#data_most_inf_feat <- predictors(data_rfe)
#data_df_inf_num <- subset(data_df,select=)
```

```{r data_df_descr, include=FALSE, echo=FALSE}

data_n_obs <- length(data_df[,1])
data_n_resp_levels <- length(levels(data_df[,data_response_index]))
data_resp_levels <- levels(data_df[,data_response_index])
data_resp_levels_string <- paste(paste0(data_resp_levels[1:(data_n_resp_levels-1)], collapse=", "), tail(data_resp_levels, n=1), sep=" and ")


```


```{r dyagnosis_function, include=FALSE, echo=FALSE}
factor_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

dyagnose_data <- function(df){
  data_dyagnosis <- data.frame(stringsAsFactors=FALSE)
  data_length <- length(df[1,])
  for (i in 1:data_length) {
    rownames(data_dyagnosis[i,]) <- colnames(df[i])
    data_dyagnosis$class[i] <- class(df[,i])
    if (class(df[,i]) == "numeric"){
      data_descr <- describe(data_df[,i])
      data_dyagnosis$class[i] <- class(df[,i])
      data_dyagnosis$n_levels[i] <- "NA"
      data_dyagnosis$mean[i] <- data_descr$mean
      data_dyagnosis$median[i] <- data_descr$median
      data_dyagnosis$sd[i] <- data_descr$sd
      data_dyagnosis$min[i] <- data_descr$min
      data_dyagnosis$max[i] <- data_descr$max
      data_dyagnosis$range[i] <- data_descr$range
      data_dyagnosis$skew[i] <- data_descr$skew
      data_dyagnosis$kurtosis[i] <- data_descr$kurtosis
      data_dyagnosis$se[i] <- data_descr$se
#      normality <- shapiro.test(df[,i])
      normality <- ks.test(df[,i],y='pnorm', mean=1, sd=2)
      data_dyagnosis$normality[i] <- ifelse(normality$p.value >= 0.05,1,0)

      data_dyagnosis$var_zero[i] <- ifelse(var(data_df[,i]) == 0,1,0)
      data_dyagnosis$description[i] <- paste("The **", rownames(data_dyagnosis[i,]), "** variable is a numeric variable with the following descriptive statistics: \n**mean** = ", data_dyagnosis$mean[i], ", \n**median** = ", data_dyagnosis$median[i], ", \n**standard deviation** = ", data_dyagnosis$sd[i], ", \n**min value** = ", data_dyagnosis$min[i], ", \n**max value** = ", data_dyagnosis$max[i], ", \nwhich accounts for **range** = ", data_dyagnosis$range[i], ". \nIt has a **skewness** of ", data_dyagnosis$skew[i], ", a **kurtosis** of ", data_dyagnosis$kurtosis[i], ", and a **standard error** of ", data_dyagnosis$se[i], ".\n\n", sep = "")
    } else if (class(df[,i]) == "factor"){
      data_dyagnosis$class[i] <- class(df[,i])
      data_dyagnosis_levels <- levels(df[,i])
      data_dyagnosis$n_levels[i] <- length(data_dyagnosis_levels)
      data_dyagnosis$mean[i] <- "NA"
      data_dyagnosis$median[i] <- "NA"
      data_dyagnosis$sd[i] <- "NA"
      data_dyagnosis$min[i] <- "NA"
      data_dyagnosis$max[i] <- "NA"
      data_dyagnosis$range[i] <- "NA"
      data_dyagnosis$skew[i] <- "NA"
      data_dyagnosis$kurtosis[i] <- "NA"
      data_dyagnosis$se[i] <- "NA"
      data_dyagnosis$normality[i] <- "NA"
      data_dyagnosis$var_zero[i] <- "NA"
      if (data_dyagnosis$n_levels[i] > 1){
        if (data_dyagnosis$n_levels[i] < 5) {
          data_dyagnosis$description[i] <- paste("The **", rownames(data_dyagnosis[i,]), "** variable is a factor variable with ", data_dyagnosis$n_levels[i], " levels, being these ", paste(paste0(data_dyagnosis_levels[1:(data_dyagnosis$n_levels[i]-1)], collapse=", "), tail(data_dyagnosis_levels, n=1), sep=" and "), ". The mode is ", factor_mode(df[,i]), " with ", table(df[,i])[names(table(df[,i])) == factor_mode(df[,i])]," occurrences.\n\n", sep = "")
        } else {
          data_dyagnosis$description[i] <- paste("The **", rownames(data_dyagnosis[i,]), "** variable is a factor variable with ", data_dyagnosis$n_levels[i], " levels, being these ", paste(data_dyagnosis_levels[1:4], sep=", "), " and others (multiple or ordered levels, non-practical to enumerate). The mode is ", factor_mode(df[,i]), " with ", table(df[,i])[names(table(df[,i])) == factor_mode(df[,i])]," occurrences.\n\n", sep = "")
        }
      } else {
        data_dyagnosis$description[i] <- paste("The **", rownames(data_dyagnosis[i,]), "** variable is a factor variable with 1 level, being this ", data_dyagnosis_levels[1], ". As it's the only level for this factor, it is also the mode.", sep = "")
      }
    }
    data_dyagnosis$has_na[i] <- ifelse("FALSE" %in% complete.cases(data_df[,i]),1,0)
  }
  return(data_dyagnosis)
}

identify_cp <- function(dyag_df) {
  data_cp <- data.frame(stringsAsFactors=FALSE)
  data_length <- length(dyag_df[,1])
  for (i in 1:data_length) {
    if (dyag_df$normality[i] == 0 || dyag_df$has_na[i] == 1 || dyag_df$var_zero[i] == 1) {
      rbind(data_cp,dyag_df[i,]) 
    }
  }
  return(data_cp)
}

```
```{r dyagnosis_df, include=FALSE, echo=FALSE}
data_dyag_df <- dyagnose_data(data_df)
data_dyag_cp <- subset(data_dyag_df, data_dyag_df$has_na == 1 | data_dyag_df$var_zero == 1)
data_dyag_var_zero <- data_dyag_df[which(data_dyag_df$var_zero == 1),]
data_dyag_has_na <- data_dyag_df[which(data_dyag_df$has_na == 1),]

data_dyag_cp_has_na_vars <- if (length(data_dyag_has_na) > 1) {
  paste(paste0(head(rownames(data_dyag_has_na),-1), collapse=", "), tail(rownames(data_dyag_has_na),1), sep=" and ")
} else {
  rownames(data_dyag_has_na[,1])
}

data_dyag_cp_var_zero_vars <- if (length(data_dyag_var_zero) > 1) {
  paste(paste0(head(rownames(data_dyag_var_zero),-1), collapse=", "), tail(rownames(data_dyag_var_zero),1), sep=" and ")
} else {
  rownames(data_dyag_var_zero[,1])
}

```

#Introduction  

This document aims to present a complete descriptive statistics briefing on the specific data set it is based on. It's **not** to be a *must-read-wholly* file, as it is data-intensive and not all statistics may be needed for most variables. Even so, for the sake of complete information and critic sense, it **must** be available at the reader's discretion, in order for it to be a tool to use in judgement of anomalous results. The statistics presented in this document may be updated, extended or improved at any time it is deemed necessary and used to complement the critic appraisal of results in the wider context of this project.  

#1 - Data set general data

The current data set is a study on **normal**, **atypical**, and reactive lymphocytes, in which 2867 numerical variables, based on colorimetric and geometric features have been measured. The aim of the study is more thoroughly explained both in the proposal and main report for this project.  

The current data set has `r data_n_obs` observations for `r length(data_df[1,])` variables, of which `r data_class_factors` are factors and `r data_class_numeric` are numeric.  

#2 - Critic point variables  

Of all variables included in the data set, `r length(data_dyag_cp[,1])` are potential critical point variables, and should be reviewed as such. `r length(data_dyag_df$var_zero[which(data_dyag_df$var_zero == 1)])` have variance zero and should be considered as constants and be treated as such. `r length(data_dyag_df$has_na[which(data_dyag_df$has_na == 1)])` of `r data_class_numeric` have missing values and need a further review.  

Even though non-normality could be considered a critic point for a variable, the actual data set is composed of `r data_n_obs` observations. A data set this large creates an artifact in which Kolmogorov-Smirnoff or Shapiro-Wilk tests tend to reject the null hypothesis of normality at a false rate, so it won't be assessed in this report.

```{r data_desc_critic_point, results='asis', include=TRUE, echo=FALSE}
if (length(data_dyag_var_zero[,1]) > 0) {
  cat(paste("**Variance zero:** *", data_dyag_cp_var_zero_vars, sep = ""))
  cat("\n\n")
} else {
  cat("**Variance zero:** no variables present variance zero.")
  cat("\n\n")
}
if (length(data_dyag_has_na[,1]) > 0) {
  cat(paste("**Has NA:** *", data_dyag_cp_has_na_vars, "*\n\n", sep = ""))
  cat("\n\n")
} else {
  cat("**Has NA:** no variables present NA values.")
  cat("\n\n")
}
```

#3 - Total variable descriptive statistics  

In this section, descriptive statistics are presented for **ALL** variables. Depending on the number of variables, this can be a long, intensive section in which specific data should be searched for. 
The **response variable** *`r colnames(data_df[data_response_index])`* has `r data_n_resp_levels` levels, being these `r data_resp_levels_string`.

```{r data_desc_stats_cat, results='asis', include=TRUE, echo=FALSE}
for (i in (1:length(data_dyag_df[,1]))) {
  if (i != data_response_index) {
    cat(data_dyag_df[i,"description"])
    cat("\n\n")
  }
}
```

