---
title: "Lymphocyte Data Diagnosys Script"
author: "Giménez Gredilla, Daniel"
date: "November 2017"
bibliography: library.bib
output:
  pdf_document: 
    toc: yes
    toc_depth: 3
    fig_caption: yes
    includes:  
      in_header: tex-support.tex
options:
  encoding: utf8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.lp = "Fig. ", tidy.opts=list(width.cutoff=60),tidy=TRUE, cache = TRUE, fig.width=12, fig.height=4, comment = NULL, out.height='200px', dpi=200, fig.pos = 'h', out.extra = '')
```

```{r library_load, include=FALSE, echo=FALSE, warning=FALSE}
require(car)
require(psych)
require(caret)
require(mlbench)
require(randomForest)
```

```{r data_var_declare, include=FALSE, echo=FALSE}
data_class_factors <- 7
data_class_numeric <- 2867
data_response_index <- 2
```

```{r data_load, include=FALSE, echo=FALSE}
colClasses <- append(c(rep("factor",data_class_factors)),c(rep("numeric",data_class_numeric)))
data_df <- read.csv2(file = "resources/data_dummy.csv", sep = ",", dec = ".", colClasses = colClasses, stringsAsFactors = FALSE)
#data_df <- read.csv2(file = "resources/data.csv", sep = ",", dec = ".", colClasses = colClasses, stringsAsFactors = FALSE)
data_formula <- as.formula(paste(colnames(data_df[data_response_index]), "~", paste(colnames(data_df)[-(1:data_class_factors)], collapse=" + ")))
```

```{r data_df_str_summ, include=FALSE, echo=FALSE}
data_descr <- describe(data_df)
data_n_obs <- data_descr[1,]$n
data_n_resp_levels <- length(levels(data_df[,data_response_index]))
data_resp_levels <- levels(data_df[,data_response_index])
data_resp_levels_string <- paste(paste0(data_resp_levels[1:(data_n_resp_levels-1)], collapse=", "), tail(data_resp_levels, n=1), sep=" and ")
data_control <- rfeControl(functions=rfFuncs, method="cv", number=10)
data_rfe <- rfe(data_df[,(data_class_factors+1):length(data_df[1,])], data_df[,data_response_index], sizes = c(1:20), metric = "Accuracy", maximize = TRUE, rfeControl = data_control)
data_most_inf_feat <- predictors(data_rfe)
```

```{r dyagnosis_function, include=FALSE, echo=FALSE}
factor_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

dyagnose_data <- function(df){
  data_dyagnosis <- data.frame(stringsAsFactors=FALSE)
  data_length <- length(df[1,])
  for (i in 1:data_length) {
    rownames(data_dyagnosis[i,]) <- colnames(df[i])
    data_dyagnosis$class[i] <- class(df[,i])
    if (class(df[,i]) == "numeric"){
      data_dyagnosis$class[i] <- class(df[,i])
      data_dyagnosis$n_levels[i] <- "NA"
      data_dyagnosis$mean[i] <- data_descr[i,]$mean
      data_dyagnosis$median[i] <- data_descr[i,]$median
      data_dyagnosis$sd[i] <- data_descr[i,]$sd
      data_dyagnosis$min[i] <- data_descr[i,]$min
      data_dyagnosis$max[i] <- data_descr[i,]$max
      data_dyagnosis$range[i] <- data_descr[i,]$range
      data_dyagnosis$skew[i] <- data_descr[i,]$skew
      data_dyagnosis$kurtosis[i] <- data_descr[i,]$kurtosis
      data_dyagnosis$se[i] <- data_descr[i,]$se
      normality <- shapiro.test(df[,i])
      homocedasticity <- shapiro.test(df[,i])
      data_dyagnosis$normality[i] <- ifelse(normality$p.value >= 0.05,1,0)
      data_dyagnosis$var_zero[i] <- ifelse(var(data_df[,i]) == 0,1,0)
      data_dyagnosis$description[i] <- paste("The ", rownames(data_dyagnosis[i,]), " variable is a numeric variable with the following descriptive statistics: mean = ", data_dyagnosis$mean[i], ", median = ", data_dyagnosis$median[i], ", standard deviation = ", data_dyagnosis$sd[i], ", min value = ", data_dyagnosis$min[i], ", max value = ", data_dyagnosis$max[i], ", which accounts for range = ", data_dyagnosis$range[i], ". It has a skew of ", data_dyagnosis$skew[i], ", a kurtosis of ", data_dyagnosis$kurtosis[i], ", and a standard error of ", data_dyagnosis$se[i], ".", sep = "")
    } else if (class(df[,i]) == "factor"){
      data_dyagnosis$class[i] <- class(df[,i])
      data_dyagnosis_levels <- levels(df[,i])
      data_dyagnosis$n_levels[i] <- length(data_dyagnosis_levels)
      data_dyagnosis$mean[i] <- "NA"
      data_dyagnosis$median[i] <- "NA"
      data_dyagnosis$sd[i] <- "NA"
      data_dyagnosis$min[i] <- "NA"
      data_dyagnosis$max[i] <- "NA"
      data_dyagnosis$range[i] <- "NA"
      data_dyagnosis$skew[i] <- "NA"
      data_dyagnosis$kurtosis[i] <- "NA"
      data_dyagnosis$se[i] <- "NA"
      data_dyagnosis$normality[i] <- "NA"
      data_dyagnosis$var_zero[i] <- "NA"
      if (data_dyagnosis$n_levels[i] > 1){
        if (data_dyagnosis$n_levels[i] < 5) {
          data_dyagnosis$description[i] <- paste("The **", rownames(data_dyagnosis[i,]), "** variable is a factor variable with ", data_dyagnosis$n_levels[i], " levels, being these ", paste(paste0(data_dyagnosis_levels[1:(data_dyagnosis$n_levels[i]-1)], collapse=", "), tail(data_dyagnosis_levels, n=1), sep=" and "), ". The mode is ", factor_mode(df[,i]), " with ", table(df[,i])[names(table(df[,i])) == factor_mode(df[,i])]," occurrences.", sep = "")
        } else {
          data_dyagnosis$description[i] <- paste("The ", rownames(data_dyagnosis[i,]), " variable is a factor variable with ", data_dyagnosis$n_levels[i], " levels, being these ", paste(data_dyagnosis_levels[1:4], sep=", "), " and others (multiple or ordered levels, non-practical to enumerate). The mode is ", factor_mode(df[,i]), " with ", table(df[,i])[names(table(df[,i])) == factor_mode(df[,i])]," occurrences.", sep = "")
        }
      } else {
        data_dyagnosis$description[i] <- paste("The **", rownames(data_dyagnosis[i,]), "** variable is a factor variable with 1 level, being this ", data_dyagnosis_levels[1], ". As it's the only level for this factor, it is also the mode.", sep = "")
      }
    }
    data_dyagnosis$has_na[i] <- ifelse("FALSE" %in% complete.cases(data_df[,i]),1,0)
  }
  return(data_dyagnosis)
}

identify_cp <- function(dyag_df) {
  data_cp <- data.frame(stringsAsFactors=FALSE)
  data_length <- length(dyag_df[,1])
  for (i in 1:data_length) {
    if (dyag_df$normality[i] == 0 || dyag_df$has_na[i] == 1 || dyag_df$var_zero[i] == 1) {
      rbind(data_cp,dyag_df[i,]) 
    }
  }
  return(data_cp)
}

```
```{r dyagnosis_df, include=FALSE, echo=FALSE}
data_dyag_df <- dyagnose_data(data_df)
data_dyag_cp <- subset(data_dyag_df, data_dyag_df$normality == 0 | data_dyag_df$has_na == 1 | data_dyag_df$var_zero == 1)
data_dyag_non_normal <- data_dyag_df[which(data_dyag_df$normality == 0),]
data_dyag_var_zero <- data_dyag_df[which(data_dyag_df$var_zero == 1),]
data_dyag_has_na <- data_dyag_df[which(data_dyag_df$has_na == 1),]

data_dyag_cp_non_normal_vars <- if (length(data_dyag_non_normal) > 1) {
  paste(paste0(head(rownames(data_dyag_non_normal),-1), collapse=", "), tail(rownames(data_dyag_non_normal),1), sep=" and ")
} else {
  rownames(data_dyag_non_normal[,1])
}

data_dyag_cp_has_na_vars <- if (length(data_dyag_has_na) > 1) {
  paste(paste0(head(rownames(data_dyag_has_na),-1), collapse=", "), tail(rownames(data_dyag_has_na),1), sep=" and ")
} else {
  rownames(data_dyag_has_na[,1])
}

data_dyag_cp_var_zero_vars <- if (length(data_dyag_var_zero) > 1) {
  paste(paste0(head(rownames(data_dyag_var_zero),-1), collapse=", "), tail(rownames(data_dyag_var_zero),1), sep=" and ")
} else {
  rownames(data_dyag_var_zero[,1])
}

```

#Introduction  

This document aims to present a complete descriptive statistics briefing on the specific data set it is based on. It's **not** to be a *must-read-wholly* file, as it is data-intensive and not all statistics may be needed for most variables. Even so, for the sake of complete information and critic sense, it is presented for the reader's discretion. The statistics presented in this document may be updated at any time it is deemed necessary and used to complement the critic appraisal of results in the wider context of this project.  

#1 - Data set general data

The current data set is a study on **normal**, **atypical**, and reactive lymphocytes, in which 2867 numerical variables, based on colorimetric and geometric features have been measured. The aim of the study is more thoroughly explained both in the proposal and main report for this project.  

The current data set has `r data_n_obs` observations for `r length(data_descr[,1])` variables, of which `r data_class_factors` are factors and `r data_class_numeric` are numeric.  

  

#2 - Critic point variables  

Of all variables included in the data set, `r length(data_dyag_cp[,1])` are potential critical point variables, and should be reviewed as such. `r length(data_dyag_df$var_zero[which(data_dyag_df$var_zero == 1)])` have variance zero and should be considered as constants and be treated as such. `r length(data_dyag_df$normality[which(data_dyag_df$normality == 0)])` of `r data_class_numeric` have a non-normal distribution. `r length(data_dyag_df$has_na[which(data_dyag_df$has_na == 1)])` of `r data_class_numeric` have missing values and need a further review.

```{r data_desc_critic_point, results='asis', include=TRUE, echo=FALSE}
if (length(data_dyag_non_normal[,1]) > 0) {
  cat(paste("**Non-normal:** *", data_dyag_cp_non_normal_vars, sep = ""))
  cat("\n\n")
}
if (length(data_dyag_var_zero[,1]) > 0) {
  cat(paste("**Variance zero:** *", data_dyag_cp_var_zero_vars, sep = ""))
  cat("\n\n")
}
if (length(data_dyag_has_na[,1]) > 0) {
  cat(paste("**Has NA:** *", data_dyag_cp_has_na_vars, "*\n\n", sep = ""))
  cat("\n\n")
}
```

#3 - Most influential variable descriptive statistics

By fitting a simple, raw **Random Forest** algorithm, a small set of potentially most accurate variables for classification is subset. Descriptive statistics are listed for these features. It is important to note that these are just **potentially** most accurate variables, **entirely subject** to later dimension reduction techniques on a more informed context.  

```{r data_most_inf_feat, results='asis', include=TRUE, echo=FALSE}
data_most_inf_feat_string <- paste(paste0(data_most_inf_feat[1:(length(data_most_inf_feat)-1)], collapse=", "), tail(data_most_inf_feat, n=1), sep=" and ")
data_dyag_most_inf <- data_dyag_df[which(rownames(data_dyag_df) %in% data_most_inf_feat),]
```

According to this simple, raw **RF**, the most influential variables are `r data_most_inf_feat_string`, and their main descriptive statistics are as follows:  

```{r data_most_inf_feat_desc, results='asis', include=TRUE, echo=FALSE}
for (i in (1:length(data_dyag_df[,1]))) {
  if (i != data_response_index) {
    cat(data_dyag_df[i,"description"])
    cat("\n\n")
  }
}
```

#4 - Total variable descriptive statistics  

In this section, descriptive statistics are presented for **ALL** variables. Depending on the number of variables, this can be a long, intensive section in which specific data should be searched for.  

The **response variable** *`r colnames(data_df[data_response_index])`* has `r data_n_resp_levels` levels, being these `r data_resp_levels_string`.

```{r data_desc_stats_cat, results='asis', include=TRUE, echo=FALSE}
for (i in (1:length(data_dyag_df[,1]))) {
  if (i != data_response_index) {
    cat(data_dyag_df[i,"description"])
    cat("\n\n")
  }
}
```

